Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-03-19 manuscript.tex
Section figure 7 done!

\subsection*{
  Figure 6. Deciphering the enhancer code of early neuronal differentiation.
}

@@

@@

\subsection*{Figure 7. }

>>>>

\subsection* {
  Figure 6. Deciphering the enhancer code of early neuronal differentiation.
}

@@

\subsection* {
  Four to five distinct transcription factor binding sites in a region of around 240 base pairs is sufficient to encode
  all cell states related to neuronal progenitors, early neuronal differentiation and neural crest development.
}

Summarizing our results, we identified a total of sixteen and fifteen distinct cell states in respectively the neural
tube organoids and four weeks post conception embryo head with almost one-to-one correspondence based on TFBS importances and TF expression. Additionally, we identified 4 distinct cell states corresponding to AP patterning of the neuronal progenitors in the embryo. To get insight into the minimal set of TFBS that are needed to encode these cell states we trained logistic regression models to classify genomic regions as to belonging to the topic representing each state or not, making use of the motif scores of the identified patterns from each code table (\textbf{Fig. 3e}, \textbf{Fig. 4e}, \textbf{Fig. 5b} and \textbf{Fig. 6b}). The overall performance of these models was acceptably high (\textbf{Fig. 7a}) with an average area under the ROC of X and area under the precision-recall curve of X. Notably, all models for neural crest related states performed best (\textbf{Fig. 7a}). Next, we selected patterns that had a positive coefficient for at least one of the logistic regression models resulting in a total of twenty one patterns for fifteen to sixteen states plus the additional four AP states (\textbf{Fig. 7a}). Finally, we used the identified patterns to design synthetic enhancers for each state using motif embedding\cite{taskiran2024synthetic, kempynck2025crested}. We used at least three TFBS in the design process for each state. In case less than three distinct patterns had a positive coefficient for a state, we implanted multiple copies of each pattern to come to a total of at least four TFBS. The resulting synthetic DNA sequences had an overall high prediction score (X.X on average) (\textbf{Fig. 7a}) - and were cell type specific (\textbf{Fig. 7b}) - with the prediction score for neural crest related states being again the highest. Note that some TFBS like FOX and SOX are used by many cell states (\textbf{Fig. 7a}) while other are more specific to certain states. \par

To conclude, enhancers specific to DV-patterning, neural crest states and early neuronal differentiation tend to be encoded in
the human genome using around four to five distinct TFBS (\textbf{Fig. 7a}) in regions of on average 240-243bp +/- 67-77bp in size (\textbf{Fig. 7c}). AP-patterning enhancers seem to be an exception to this rule, where they are encoded using only two and sometimes even a single TFBS (\textbf{Fig. 7a}).

@@

\subsection*{
  Figure 7. Neural tube cell states are encoded using four to five distinct transcription factor binding
  sites
}.
\textbf{a}, Area under the receiver-operator curve (outlined dots) and precision-recall curve of logistic regression
models classifying genomic regions to each topic using motif scores as features (top) and patterns with positive
coefficient for each logistic regression model along with the number of each patterns used for enhancer design
represented by the size of each dot (middle) and boxplots (n = 200) showing prediction score of designed enhancers for
each topic (bottom). \textbf{b}, heatmap showing prediction score for each topic of enhancers designed for all topics
for using $DeepNeuralTube_{organoid}$ (left) and $DeepNeuralTube_{embryo}$ (right). \textbf{c}, scatter plot showing
location of all pattern instances and contribution score at those locations, color is -log p value of the motif score
for each pattern calculated using FIMO\cite{grant2011fimo, schreiber2025tangermeme} for organoid (top left) and embryo
(top right) and histograms showing the number of motif instances per 25 bp bins with confidence intervals indicated in
gray scales (black is +/- 1 standard deviation, gray is +/- 2 standard deviation and white is more than 3 standard
deviations) for organoid (bottom left) and embryo (bottom right).
